---
title: <center><b>CSE3505 - Essentials of Data Analytics - G2<br>Review - 2</b></center>
output: html_document
---
<pre>
<b> Name: Soumik Kabiraj - 20BCE1504</b>
<b>       Panav Sinha - 20BCE1640</b>
<b>       Mayank Yadav - 20BCE1674</b>
</pre>
<br><br>

```{r}
library(dplyr)
library(tidyverse)
library(caret)
library(grid)
library(gridExtra)
library(mlbench)
library(caretEnsemble)
library(ggrepel)
library(reshape2)
library(ggExtra)
library(GGally)
library(hrbrthemes)
library(ROSE)
```

```{r}
df <- read.csv("smoking.csv")
summary(df)
str(df)
```

```{r}
#Tartar is the hard calcified deposits that form and coat the teeth and gums
unique(df$tartar)
unique(df$gender)
unique(df$oral)
df$sex_num <- ifelse(df$gender=="F",0,1)
df$tartar <-  ifelse(df$tartar=="Y",1,0)
class(df$oral)
df$oral <- as.numeric(as.factor(df$oral))
class(df$oral)
```

<h3> Are there outliers?
#### Outliers for continuous variables
```{r}
df_num<-select_if(df,is.numeric) %>% select(-c(ID,hearing.right.,hearing.left.,smoking,dental.caries,Urine.protein))
head(df_num)
df_num_p<-df_num %>% gather(variable,values,1:18)
head(df_num_p)

options(repr.plot.width = 18, repr.plot.height = 14)
ggplot(df_num_p)+geom_boxplot(aes(x=variable,y=values),fill="royalblue") + facet_wrap(~variable,ncol=3,scales="free") + theme(strip.text.x = element_blank(),text = element_text(size=8))
```

```{r}
a <- ggplot(df, aes(age))+
  geom_boxplot(fill="steelblue")+scale_x_continuous(breaks=seq(0,100,5))+labs(y="Age")+coord_flip()

b <- ggplot(df, aes(eyesight.left.))+
  geom_boxplot(fill="steelblue")+labs(y="Eyesight left")+coord_flip()

c <- ggplot(df, aes(eyesight.right.))+
  geom_boxplot(fill="steelblue")+labs(y="Eyesight right")+coord_flip()

d <- ggplot(df, aes(HDL))+
  geom_boxplot(fill="steelblue")+labs(y="HDL")+coord_flip()

e <- ggplot(df, aes(height.cm.))+
  geom_boxplot(fill="steelblue")+labs(y="Height")+coord_flip()

f <- ggplot(df, aes(triglyceride))+
  geom_boxplot(fill="steelblue")+labs(y="Triglyceride")+coord_flip()

g <- ggplot(df, aes(ALT))+
  geom_boxplot(fill="steelblue")+labs(y="ALT")+coord_flip()

options(repr.plot.width = 18, repr.plot.height = 12)
grid.arrange(a,b,c,d,e,f,g, ncol=4)
```


```{r}
df$age_grp <- cut(df$age, c(0,17,60,100,120), labels = c("0-17","18-60","60-100","100+") )
head(df$age_grp)

h <- ggplot(df,aes(x=`age_grp`, fill=`gender`))+geom_bar()+facet_grid(.~gender)+
  stat_count(aes(y=..count.., label=..count..), vjust=-0.5,geom="text", col="black", size=3.5)+
  labs(x="Age Group", y = "Count", title="Age Group vs Sex", fill= "Sex")+
  theme(plot.title=element_text(face="bold",  hjust=0.5), legend.position = "bottom",text=element_text(size=10))+
    scale_fill_manual(values=c("plum3","royalblue"))

i <- ggplot(df, aes(y = gender,fill=gender))+
  geom_bar()+
  scale_fill_manual(values=c("plum3","royalblue"))+
  labs(title="Sex",x=" ")+
  theme(legend.position = "none", plot.title=element_text(face="bold",  hjust=0.5), text=element_text(size=10))+
  coord_polar("y")

options(repr.plot.width = 16, repr.plot.height = 12)
grid.arrange(i, h, ncol=1)
```

```{r}
df$bmi <- (df$weight.kg./(df$height.cm.*df$height.cm.))*10000
df$bmi_cat <- ifelse(df$bmi<18.5,"Underweight", NA )
df$bmi_cat <- ifelse(df$bmi>=18.5 & df$bmi <=24.9,"Healthy weight", df$bmi_cat )
df$bmi_cat <- ifelse(df$bmi > 24.9 & df$bmi<=29.9,"Overweight", df$bmi_cat)
df$bmi_cat <- ifelse(df$bmi > 30,"Obesity", df$bmi_cat)
table(df$bmi_cat)
```

```{r}
ggplot(df, aes(x=bmi_cat, fill= as.factor(smoking)))+
  geom_bar(position="dodge", col="black")+
  scale_fill_manual(label=c("no","yes"),values=c("cadetblue", "firebrick"))+
  theme(legend.position = "bottom")+
  labs(x=" ", fill="Smoking", title="BMI and Smoking")+
  theme(text=element_text(size=18), plot.title= element_text(face="bold", hjust=0.5))
```

```{r}
msel <- df%>%
  dplyr::filter(smoking==1)%>%
  summarize(mean(eyesight.left.))
sprintf("Mean left eyepower of smokers: %f",msel)

mnsel <- df%>%
  dplyr::filter(smoking==0)%>%
  summarize(mean(eyesight.left.))
sprintf("Mean left eyepower of non-smokers: %f",mnsel)

mser <- df%>%
  dplyr::filter(smoking==1)%>%
  summarize(mean(eyesight.right.))
sprintf("Mean right eyepower of smokers: %f",mser)

mnser <- df%>%
  dplyr::filter(smoking==0)%>%
  summarize(mean(eyesight.right.))
sprintf("Mean right eyepower of non-smokers: %f",mnser)
```

```{r}
BP <- c("Normal","Elevated","Hypertension I","Hypertension II", "Hypertensive Crisis")
sys <- c("less than 120", "120-129", "130-139", "140 or higher", "more than 180")
dias <- c("less than 80", " less than 80", "80-89", "90 or higher", "120 or higher")
pressure <- data.frame(BP, sys,dias)
colnames(pressure) <- c("Blood Pressure", "Systolic", "Diastolic (Relaxed)")
pressure
```

```{r}
df$pressure <- ifelse(df$systolic<120 & df$relaxation<80,"Normal",NA)
df$pressure <- ifelse(df$systolic>=120 & df$relaxation <80, "Elevated",df$pressure)
df$pressure <- ifelse(df$systolic>=130 | df$relaxation>=80,"Hypertension Stage 1",df$pressure)
df$pressure <- ifelse(df$systolic>=140 | df$relaxation>=90,"Hypertension Stage 2",df$pressure)
df$pressure <- ifelse(df$systolic>=180 | df$relaxation>=120,"Hypertensive Crisis",df$pressure)
table(df$pressure)
```

```{r}
table <- df %>% 
  group_by(df$pressure)%>%
  summarise(smoker= length(smoking[smoking=="1"]), 
            nonsmoker= length(smoking[smoking=="0"]),
            smoker_percent= smoker/sum(smoker,nonsmoker)*100)%>%
  arrange(desc(smoker_percent))

table
```

```{r}
options(repr.plot.width = 16, repr.plot.height = 12)
ggplot(df, aes(x=reorder(pressure,pressure,function(x)-length(x)), fill=as.factor(smoking)))+
  geom_bar(col="black",position="stack")+
  theme(plot.title=element_text(face="bold",hjust=0.5),legend.position = "bottom", axis.text.x = element_text(angle=30), text=element_text(size=18))+
  labs(title="Blood Pressure and Smoking",x="Blood Pressure", fill="Smoking")+
  stat_count(aes(y=..count..,label=..count..), geom="text", vjust=1.5, size=8, col="white")+
  scale_fill_manual(label=c("no","yes"),values=c("cadetblue", "firebrick"))
```

<h3> Does smoking increase cholesterol?
#### Cholesterol: Do smokers have lower good fat HDL? Does smoking increase harmful fats?
```{r}
options(repr.plot.width = 16, repr.plot.height = 12)
ggplot(subset(df,df$HDL<200), aes(x=HDL, fill=as.factor(smoking)))+
  geom_bar()+
  labs(title="HDL and Smoking", fill="Smoking")+
  theme(legend.position = "bottom", plot.title = element_text(face="bold", hjust=0.5), text=element_text(size=18))+
  scale_x_continuous(breaks=seq(0,200,20))+
  scale_fill_manual(labels=c("No","Yes"),values=c("forest green","red"))
```

<h3> LDL (the bad cholesterol)

```{r}
options(repr.plot.width = 16, repr.plot.height = 12)
ggplot(subset(df,df$LDL<200), aes(x=LDL, fill=as.factor(smoking)))+
  geom_bar()+
  labs(title="LDL and Smoking", fill="Smoking")+
  theme(legend.position = "bottom", plot.title = element_text(face="bold", hjust=0.5), text=element_text(size=18))+
  scale_x_continuous(breaks=seq(0,200,20))+
  scale_fill_manual(labels=c("No","Yes"),values=c("forest green","red"))
```

<h3> Triglycerides
#### The cutoff for triglyceride is 150, above 150 is at risk.
```{r}
df%>% select(c("triglyceride", "smoking")) %>% 
  group_by(smoking) %>%
  summarise(tryg_mean = mean(triglyceride)) %>%
  mutate(., Type=ifelse(tryg_mean<150,"Healthy","Risk"))
```

<h3> Liver: Are liver functions worse for smokers? 
#### AST normal range - 8 to 33 U/L
#### ALT normal range - 4 to 36 U/L
#### GTP normal range - 5 to 40 U/L

```{r}
df%>% select(c("ALT", "AST", "Gtp", "smoking")) %>% 
  group_by(smoking) %>%
  summarise(Alt_mean = mean(ALT),
            Ast_mean = mean(AST),
            Gtp_mean = mean(Gtp))
```

<h3> Blood: Do smokers have higher hemoglobin?
```{r}
mean(df[df$smoking=="1",'hemoglobin'])
mean(df[df$smoking=="0",'hemoglobin'])

options(repr.plot.width = 14, repr.plot.height = 10)
ggplot(df, aes(y=hemoglobin))+
  geom_boxplot(fill=c("cadetblue","firebrick2"))+
  facet_grid(.~as.factor(smoking))+
  scale_y_continuous(breaks= seq(0,20,2))+
  labs(title="Hemoglobin and Smoking", fill="smoking")+
  theme(plot.title = element_text(face="bold", hjust=0.5), text=element_text(size=18))
```


<h3> Random Forest Model 
```{r}
# Split data
library(caTools)
set.seed(123)
df$split<- sample.split(df$smoking, SplitRatio = 0.7)
df_train <- df %>% filter( split==TRUE) %>% select(-split)
df_test <- df%>% filter(split==FALSE)%>%select(-split)
 
paste0("Number of rows in train dataset: ",nrow(df_train))
paste0("Number of rows in test dataset: ", nrow(df_test))
```

```{r}
library(randomForest)
set.seed(1234)

model1 <- randomForest(factor(smoking) ~ age + height.cm. + weight.kg.+ sex_num + 
                         eyesight.left. + eyesight.right. + hearing.left. + hearing.right. +
                         systolic + relaxation + 
                         Cholesterol + triglyceride + HDL + LDL +
                         fasting.blood.sugar + hemoglobin + Urine.protein + serum.creatinine+
                         AST + ALT + Gtp +
                         dental.caries + tartar,
                                importance=TRUE,
                                ntree= 500,
                                data = df_train)
```

```{r}
options(repr.plot.width = 14, repr.plot.height = 10)
plot(model1, ylim=c(0,0.36),main="Error Plot")
legend('topright', colnames(model1$err.rate), col=1:3, fill=1:3)
```

<h2> Plot important variables.
```{r}
imp <- importance(model1)

# Creating a dataframe of variable importance metrics
df_imp<- data.frame(variable=row.names(imp),
                     importance = imp[,'MeanDecreaseGini'],
                     accuracy= imp[,'MeanDecreaseAccuracy'])
noquote("Variable Importance Dataframe")
df_imp

# Visualizations of importance variables
options(repr.plot.width = 14, repr.plot.height = 10)
ggplot(df_imp, aes(x=reorder(variable,-importance),y=importance, fill=importance))+
  geom_col()+
  labs(x="Predictors", y ="Importance", title="Importance of predictor variables")+
  theme(plot.title=element_text(face="bold",hjust=0.5), legend.position = "bottom", axis.text.x = element_text(angle=45), text=element_text(size=18))+
  scale_fill_continuous(type="viridis")
```


<h2> Let us remove some of the low-contribution variables and improve the model.
```{r}
model2 <- randomForest(factor(smoking) ~ age + height.cm. + weight.kg.+ sex_num + 
                         systolic + relaxation + 
                         Cholesterol + triglyceride + HDL + LDL +
                         fasting.blood.sugar + hemoglobin + serum.creatinine+
                         AST + ALT + Gtp +
                         dental.caries ,
                                importance=TRUE,
                                ntree= 500,
                                data = df_train)
paste("done")
model2
```

```{r}
model1

cat(paste("Accuracy of model 1: ", (100-17.44), "%"))

cat(paste("Accuracy of model 2: ", (100-17.76),"%"))
```

Prediction and Submission:<br>
Model 1 has higher accuracy than model 2. Hence, for further use, we will go ahead with model 1 for prediction.

```{r}
pred <- predict(model1, newdata =  df_test)
result <- data.frame(id= df_test$ID ,smoking = df_test$smoking)
result[1:20,]
```

Conclusion: <br>

Smoking is injurious to health. Here are some observations supporting this statement:

Higher eye power has been observed among smokers <br>
Hyper-intensive and high blood pressure tend to be correlated with smokers. This can cause heart diseases. <br>
The good cholesterol is seen less in smokers. <br>
Triglyceride is seen at a risk level among smokers. <br>
Gtp, a metric of liver function is seen to be at a risk level among smokers. <br>


```{r}
predict1 = predict(model1, df_test)
predict2 = predict(model2, df_test)
```

```{r}
acc1 <- mean(predict1 != df_test$smoking)
acc2 <- mean(predict2 != df_test$smoking)
cat(paste("Accuracy of model 1:", (1-acc1)*100, "%\n"))
cat(paste("Accuracy of model 2:", (1-acc2)*100, "%"))
```

```{r}
set.seed(100)
control1 <- trainControl(sampling="rose",method="repeatedcv", number=5, repeats=5)
bagCART_model <- train(factor(smoking) ~ age + height.cm. + weight.kg.+ sex_num + 
                         systolic + relaxation + 
                         Cholesterol + triglyceride + HDL + LDL +
                         fasting.blood.sugar + hemoglobin + serum.creatinine+
                         AST + ALT + Gtp +
                         dental.caries, data=df_train, method="treebag", metric="Accuracy", trControl=control1)
#Predictions on the test set
predictTest = predict(bagCART_model, df_test)
```

```{r}
acc3 <- mean(predictTest != df_test$smoking)
cat(paste("Accuracy of model 3:", (1-acc3)*100, "%"))
```


<h3>Importing the dataset</h3>
```{r}
data <- read.csv("Medicalpremium.csv")
head(data)
str(data)
dim(data)
ggcorr(data, label = T, color = "black", size = 3)+
  labs(title = "Correlation Matrix")
```
<h3>Factorizing various columns</h3>
```{r}
data$Diabetes <- as.factor(data$Diabetes)
data$BloodPressureProblems <- as.factor(data$BloodPressureProblems)
data$AnyTransplants <- as.factor(data$AnyTransplants)
data$AnyChronicDiseases <- as.factor(data$AnyChronicDiseases)
data$KnownAllergies <- as.factor(data$KnownAllergies)
data$HistoryOfCancerInFamily <- as.factor(data$HistoryOfCancerInFamily)
data$NumberOfMajorSurgeries <- as.factor(data$NumberOfMajorSurgeries)
str(data)
```
<h3>Calculating BMI</h3>
```{r}
data$bmi <- 10000*(data$Weight/(data$Height)^2)
```
<h3>Assigning categories to different BMI ranges</h3>
```{r}
data <- data %>%
  mutate( bmiCategory = case_when(
                                  bmi<18.49999 ~ "under weight",
                                  bmi>18.5 & bmi<24.99999 ~ "normal weight",
                                  bmi>25 & bmi<29.99999 ~ "over weight",
                                  bmi>30 ~ "obesity"
                                  ))
```
<h3>Diabetics People Premium Analysis</h3>
```{r}
data %>% 
  select(Diabetes,PremiumPrice) %>%
  group_by(Diabetes) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(Diabetes,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for Diabetics People")
```
<h3>Distribution of Premium Prices for Diabetic and Non-Diabetic People</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = Diabetes), color = NA, alpha = 0.6)+
  labs(title = "Density plot for Diabetics and Non-diabetic people")
```
<h3>Blood Pressure Patients Premium Analysis</h3>
```{r}
data %>% 
  select(BloodPressureProblems,PremiumPrice) %>%
  group_by(BloodPressureProblems) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(BloodPressureProblems,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for people with problem of blood pressure")
```
<h3>Distribution of Premium Prices for people with and without blood pressure problems</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = BloodPressureProblems), color = NA, alpha = 0.6)+
  labs(title = "Density plot for people with and without blood pressure problems")
```
<h3>People Gone Through Any Transplants Premium Analysis</h3>
```{r}
data %>% 
  select(AnyTransplants,PremiumPrice) %>%
  group_by(AnyTransplants) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(AnyTransplants,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for people gone through any transplants")
```
<h3>Distribution of Premium Prices for People Gone Through Any Transplants vs Those Who haven't Gone Through Any Transplants</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = AnyTransplants), color = NA, alpha = 0.6)+
  labs(title = "Density plot for people gone through any transpalnts")
```
<h3>People With Chronic Disease Premium Analysis</h3>
```{r}
data %>% 
  select(AnyChronicDiseases,PremiumPrice) %>%
  group_by(AnyChronicDiseases) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(AnyChronicDiseases,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for people with cronic disease")
```
<h3>Distribution of Premium Prices for People With Chronic Disease and People With No Chronic Disease</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = AnyChronicDiseases), color = NA, alpha = 0.6)+
  labs(title = "Density plot for  having chronic diseases")
```
<h2>Allergy Patients Premium Analysis</h2>
<h3>Average Difference in Premium Prices for Allergy patients and No Allergy Patients</h3>
```{r}
data %>% 
  select(KnownAllergies,PremiumPrice) %>%
  group_by(KnownAllergies) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(KnownAllergies,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for people having allergies")
```
<h3>Distribution of Premium Prices for Allergy Patients and No Allergy Patients</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = KnownAllergies), color = NA, alpha = 0.6)+
  labs(title = "Density plot for people with and without allergies")
```
<h2>Patients with History of Cancer in Family Premium Analysis</h2>
<h3>Average Difference in Premium Prices for Patients with History of Cancer and Patients without History of Cancer</h3>
```{r}
data %>% 
  select(HistoryOfCancerInFamily,PremiumPrice) %>%
  group_by(HistoryOfCancerInFamily) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(HistoryOfCancerInFamily,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for people with history of cancer")
```
<h3>Distribution of Premium Prices for Patients with History of Cancer and Patients without History of Cancer</h3>
```{r}
ggplot(data, aes(PremiumPrice))+
  geom_density(aes(fill = HistoryOfCancerInFamily), color = NA, alpha = 0.6)+
  labs(title = "Density plot for people with and without history of cancer")
```
<h2>People Gone Through Major Surgeries Premium Analysis</h2>
<h3>Average Difference in Premium Prices for People gone through major surgeries</h3>
```{r}
data %>% 
  select(NumberOfMajorSurgeries,PremiumPrice) %>%
  group_by(NumberOfMajorSurgeries) %>% 
  summarise( PremiumPrice = mean(PremiumPrice)) %>% 
  ggplot(.,aes(NumberOfMajorSurgeries,PremiumPrice))+
  geom_bar(stat = "identity",width = 0.4, fill = "#56B4E9", alpha = 0.6)+
  labs(title = "Bar plot for major surgeries")
```
<h3>Distribution of Premium Prices for People gone through major surgeries</h3>
```{r}
ggplot(data %>% 
         select(NumberOfMajorSurgeries,PremiumPrice) %>%
         filter(!NumberOfMajorSurgeries == 3),
         aes(PremiumPrice))+
  geom_density(aes(fill = NumberOfMajorSurgeries), color = NA, alpha = 0.6)+
  labs(title = "Density plot for people with differnt number of surgeries")
```
<h3>Box Plot for Different BMI category</h3>
```{r}
data %>% 
  mutate(bmiCategory = str_to_title(bmiCategory)) %>% 
  
  ggplot(aes(bmiCategory, PremiumPrice))+
  geom_boxplot()+
  geom_jitter(aes(color = bmiCategory),alpha = 0.4)+
  labs(title = "Distribution of Premium Price Over Differnt BMI category")
```
<h2>Prediction Model</h2>
```{r}
library(randomForest)
library(caTools)
```
```{r}
data$PremiumPrice <- as.factor(data$PremiumPrice)
summary(data)
```
<h3>Spliting Data into two Subset for Training and Testing use.</h3>
```{r}
sample <- sample.split(data$PremiumPrice, SplitRatio = 0.75)
train <- subset(data, sample == TRUE)
test <- subset(data, sample == FALSE)
dim(train)
dim(test)
```
<h3>Random Forest</h3>
```{r}
set.seed(123)
rf <- randomForest(
  PremiumPrice ~ .,
  data=train
)
print(rf)
```
```{r}
pred <- predict(rf, newdata = test[-11])
cm <- table(pred,obs = test[,11])
sum <- 0
for (i in 1:24){
    for(j in 1:24){
        if(i!=j){
            sum <-sum+cm[i,j]
        }
    }
}
sum
print(paste("The Accuracy of Random Forest Model is",(243-sum)/2.43))
```
<br>The accuracy for predictions while using 75% data for training and 25% for testing has turned out to be around 92% using Random Forest model.<br>